---
title: "4. Correcting for multiple testing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Correcting for multiple testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4.5)
```

```{r setup, include = FALSE}
library(sptrends)
r <- terra::rast(
  sort(list.files(
    example_data("vhp_ndvi"),
    pattern = "\\.tif$",
    full.names = TRUE
  ))
)
```

## Why this matters

Consider a trend test applied to a single environmental time series at a
significance level of *α* = 0.05. If the null hypothesis is true, there is a 5%
probability of committing a Type I error: rejecting the null hypothesis when
no genuine trend exists.

The situation changes dramatically with gridded data, where a separate trend
test is performed for every valid raster cell. Thousands of trend tests may
therefore be evaluated simultaneously. Even if the Type I error rate remains
at 5% for each test, the total number of false positives can become large,
allowing random variation to produce apparently meaningful spatial patterns.

Three general responses are possible. The first is to ignore multiplicity and
report cells with raw *p* < *α*. This produces selective inference
([Gutiérrez-Hernández and García,
2025](https://doi.org/10.1016/j.scitotenv.2024.177832)) because the
apparently significant results are interpreted without considering the
complete family of trend tests. The second is to control the family-wise error
rate (FWER), limiting the probability of making even one false discovery
across the entire family. This provides strict control but can substantially
reduce statistical power when thousands of cells are tested. The third is to
control the false discovery rate (FDR), accepting that some false positives
may occur while limiting their expected proportion among all results declared
significant. FDR therefore provides a more flexible balance between error
control and the ability to detect genuine trends.

## What `fdr_correction()` does

`fdr_correction()` implements three procedures for controlling the false
discovery rate. [Benjamini-Hochberg (`BH`)](https://doi.org/10.1111/j.2517-6161.1995.tb02031.x)
is the standard step-up procedure and is appropriate under independence or
recognised forms of positive dependence among tests.
[Benjamini-Krieger-Yekutieli (`BKY`)](https://doi.org/10.1093/biomet/93.3.491)
is an adaptive two-stage procedure that estimates how many null hypotheses are
likely to be true and may provide greater power under comparable dependence
assumptions ([Gutiérrez-Hernández and García,
2025](https://doi.org/10.3390/math13223630)). [Benjamini-Yekutieli
(`BY`)](https://doi.org/10.1214/aos/1013699998)
controls FDR under arbitrary dependence but is generally more conservative.

The function receives a raster or vector of *p*-values and applies BH, BKY, BY
or any requested subset. For raster inputs, missing cells are preserved and
only valid *p*-values are included in the family of simultaneous tests.

## Basic workflow

```{r}
trend <- trend_test(r, method = "CMK", report = FALSE, verbose = FALSE)
fdr <- fdr_correction(
  trend$stats$p,
  method = c("BH", "BKY", "BY"),
  q = 0.05,
  report = FALSE,
  verbose = FALSE
)
fdr
summary(fdr)
```

The summary compares all three procedures. The maps below first show the raw
CMK decisions and then the results that remain significant after BKY correction.

```{r}
bky <- fdr_correction(
  trend$stats$p,
  method = "BKY",
  q = 0.05,
  report = FALSE,
  verbose = FALSE
)
```

```{r, fig.alt = "Raw uncorrected CMK significance map"}
plot(
  trend,
  which = "maps",
  panels = "significance",
  alpha = 0.05
)
```

```{r, fig.alt = "FDR-BKY significance map"}
plot(bky)
```

The threshold plot arranges all valid *p*-values from smallest to largest. Each
point represents one trend test. Points near the beginning provide stronger
evidence against the hypothesis of no trend, whereas points farther to the
right provide weaker evidence.

The BH panel applies a standard threshold based on the total number of trend
tests. The BKY panel adds an adaptive step: it estimates what proportion of the
tested cells is likely to have no genuine trend and uses that information to
refine the threshold.

If BKY estimates that almost all null hypotheses are true, its result will be
similar to BH. If it estimates that some cells contain genuine trends, its
adaptive threshold may become less restrictive and retain more significant
results, thereby increasing statistical power.

The vertical line marks the final cutoff. Every trend test before or at that
cutoff is declared significant, whereas every test after it is not. Comparing
both panels therefore shows whether the adaptive BKY procedure retains more
discoveries than BH and why.

```{r, fig.alt = "Ordered p-values and BH and BKY rejection thresholds"}
plot(fdr, which = "threshold")
```

## Understanding the results

The unadjusted *α* is appropriate for interpreting a single trend test but is
not appropriate for raster-wide inference. Applying the same raw threshold
separately to thousands of cells ignores the accumulation of Type I errors
across the complete family of trend tests. Consequently, a map based only on
raw *p* < *α* may substantially overstate the amount of evidence for spatially
distributed trends. Multiple-testing correction is therefore required when
the results are interpreted collectively.

The target *q* defines the false discovery rate that the procedure aims to
control. For example, *q* = 0.05 means that false discoveries are expected to
represent no more than 5% of all cells declared significant, on average, under
the assumptions of the selected procedure. It is not an adjusted *α*: *α*
describes the Type I error rate of an individual unadjusted test, whereas *q*
describes the expected proportion of false discoveries across the complete
set of significant results. An adjusted threshold derived by a
multiple-testing procedure is therefore neither the original unadjusted *α*
nor the target *q*.

A cell rejected after FDR correction belongs to a family of discoveries whose
expected proportion of false discoveries is controlled. It does not receive a
separate local error guarantee. For convenience, FDR-corrected maps are often
described as showing "significant pixels", by analogy with cells satisfying an
unadjusted significance level *α*. This shorthand is useful, but it does not
mean that each highlighted pixel is individually significant at level *α* or
has an error probability equal to *q*.

## Choosing the main options

| Method | Main property | Guidance |
| --- | --- | --- |
| `BH` ([Benjamini & Hochberg, 1995](https://doi.org/10.1111/j.2517-6161.1995.tb02031.x)) | FDR control under independence or positive dependence | Recommended general starting point |
| `BKY` ([Benjamini, Krieger & Yekutieli, 2006](https://doi.org/10.1093/biomet/93.3.491)) | Adaptive FDR control under independence or positive dependence | Recommended when adaptation is expected to increase power |
| `BY` ([Benjamini & Yekutieli, 2001](https://doi.org/10.1214/aos/1013699998)) | FDR control under arbitrary dependence | Usually discard; retain only when required |

BH and BKY assume that the trend tests are independent or positively
dependent. Positive spatial dependence is often expected in gridded
environmental data because neighbouring cells tend to share similar
conditions, but it should be examined rather than assumed.
`spatial_autocorrelation()` can be used to assess whether the spatial pattern
is consistent with positive dependence. BH and BKY are the recommended
procedures: BH provides a standard and reliable starting point ([Gutiérrez-Hernández
and García, 2025](https://doi.org/10.1080/2150704X.2025.2478664)), whereas BKY is
particularly attractive when its adaptive estimation is expected to increase
statistical power ([Gutiérrez-Hernández and García,
2025](https://doi.org/10.3390/math13223630)). BY accommodates arbitrary dependence but is highly
conservative and can substantially reduce power; it would therefore be
discarded in most applications and retained only when arbitrary dependence
must be explicitly accommodated. FDR correction itself is usually very fast
relative to trend testing.

## Common mistakes

- Do not call *q* an adjusted *α*; they describe different error rates.
- Do not include missing or invalid cells in the number of tests.
- Interpreting FDR rejection as local significance.
- Redefining the testing domain after inspecting the results.

## Next steps

Continue to [`vignette("g-workflow-trends")`](g-workflow-trends.html) to
combine preprocessing, trend testing, slope estimation and FDR in one
workflow.

## Further details

See `?fdr_correction`, `?fdr_bh`, `?fdr_bky` and `?fdr_by` for
assumptions, algorithms, diagnostics, limitations and references.

## References

- Benjamini, Y. and Hochberg, Y. (1995) Controlling the False
  Discovery Rate. *Journal of the Royal Statistical Society: Series
  B*, 57, 289-300. https://doi.org/10.1111/j.2517-6161.1995.tb02031.x
- Benjamini, Y., Krieger, A.M. and Yekutieli, D. (2006) Adaptive
  Linear Step-Up Procedures That Control the False Discovery Rate.
  *Biometrika*, 93(3), 491-507.
  https://doi.org/10.1093/biomet/93.3.491
- Benjamini, Y. and Yekutieli, D. (2001) The Control of the False
  Discovery Rate in Multiple Testing Under Dependency. *Annals of
  Statistics*, 29(4), 1165-1188.
  https://doi.org/10.1214/aos/1013699998
- Gutiérrez-Hernández, O. and García, L.V. (2025) The Ghost of
  Selective Inference in Spatiotemporal Trend Analysis. *Science of
  The Total Environment*, 958, 177832.
  https://doi.org/10.1016/j.scitotenv.2024.177832
- Gutiérrez-Hernández, O. and García, L.V. (2025) Implementing the
  Linear Adaptive False Discovery Rate Procedure for Spatiotemporal
  Trend Testing. *Mathematics*, 13(22), 3630.
  https://doi.org/10.3390/math13223630
- Gutiérrez-Hernández, O. and García, L.V. (2025) False Discovery Rate
  Estimation and Control in Remote Sensing. *Remote Sensing Letters*,
  16(5), 537-548. https://doi.org/10.1080/2150704X.2025.2478664
